---
title: "Research Question 1 - 28 May 2025"
format: pdf
author: "Student ID: 201901492"
toc: false
execute: 
  cache: true
number-sections: false
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

# Research Question

To what extent do the vehicle characteristics influence injury severity outcomes for cyclists and pedestrians involved in collisions with passenger cars?


# Data Collection

in this project, data that will be used are road safety statistics that provided publicly by the UK government (STATS19).
The data consist of: 1.Road Safety – Vehicles 2.Road Safety – Casualties 3.Road Safety – Collisions

Scope of Research use data period from 2019 to 2023.

Furthermore, to obtained vehicles characteristics data (weight and dimensions), it will consider to use information based on European Environment Agency (EEA), using source code in https://github.com/itsleeds/carmatch.


a.  Load Datasets from STATS19

```{r}
library("stats19")
library(tidyverse)
library("readxl")
library("MASS")
library("pscl")
library("rms")
library("brant")
library(ROCR)
library(marginaleffects)
library(ggeffects)


#define scope of research (time period)
years <- 2020:2024

#load data road safety - vehicles
vehicles_all <-data.frame()
for (year in years){
  vehicles_data <- get_stats19(year = year, type = "veh")
  vehicles_all <-bind_rows(vehicles_all, vehicles_data)
}

vehicles_selection <- vehicles_all %>%
  dplyr::select(collision_index,
                vehicle_reference,
                generic_make_model,
                vehicle_type,
                propulsion_code,
                engine_capacity_cc,
                first_point_of_impact,
                age_of_vehicle,
                age_of_driver)

#load data road safety - collisions
collisions_all <- data.frame()
for (year in years){
  collisions_data <- get_stats19(year = year, type = "collision")
  collisions_all <-bind_rows(collisions_all, collisions_data)
}

collisions_selection <- collisions_all %>%
  dplyr::select(collision_index,
                location_easting_osgr,
                location_northing_osgr,
                latitude,
                longitude,
                police_force,
                speed_limit,
                local_authority_district,
                light_conditions,
                weather_conditions,
                road_surface_conditions,
                urban_or_rural_area)

#load data road safety - casualties
casualties_all <- data.frame()
for (year in years){
  casualties_data <- get_stats19(year = year, type = "cas")
  casualties_all <-bind_rows(casualties_all, casualties_data)
}

casualties_selection <- casualties_all %>% 
  dplyr::select(collision_index,
                vehicle_reference,
                casualty_type,
                casualty_severity,
                age_of_casualty,
                pedestrian_location,)
```

b.  Load dataset vehicles spesification such as weight and dimension based on European Environment Agency (EEA), using source code in https://github.com/ITSLeeds/carmatch

```{r}
#load dataset vehicle classification and spesification
source("https://github.com/ITSLeeds/carmatch/raw/main/R/carmatch.R")
d_raw = d_raw_euco2 <- readRDS("C:/Users/ACER/Documents/d_raw_euco2.Rds")
d_raw2 = d_raw |> 
  mutate(generic_make_model = paste(Make, Commercial_name))


#column selection
veh_char <- d_raw2 %>% 
  dplyr::select(Mass_kg,
                Wheelbase,
                SteeringAxle,
                OtherAxle,
                Engine_capacity,
                Engine_power,
                Fuel_type,
                generic_make_model)

veh_char <- veh_char %>%
  distinct(Engine_capacity, Fuel_type, generic_make_model, .keep_all = TRUE) %>%
  rename(propulsion_code = Fuel_type,
         engine_capacity_cc = Engine_capacity) %>%
  mutate(propulsion_code = tolower(propulsion_code))

veh_char1 <- veh_char %>%
  group_by(generic_make_model) %>%
  slice(1) %>%  # only first row that will be used
  ungroup() %>%
  dplyr::select(-propulsion_code, -engine_capacity_cc)
```

c. load dataset from vehicle type according to Euro Ncap

```{r}
#load dataset vehicle classification and spesification after combining manually
f <- "D:/Github/crashSeverity/Data/Raw/Vehicles_class.xlsx"

category_vehicles <- read_excel(f)

```

#Data Preprocessing

```{r}
#filling missing data on generic_make model with value in the same accident index
vehicles_selection <- vehicles_selection %>%
  group_by(collision_index) %>%
  mutate(generic_make_model = ifelse(generic_make_model == "Data missing or out of range", 
                                     first(na.omit(generic_make_model)), 
                                     generic_make_model)) %>%
  ungroup() %>%
  mutate(propulsion_code = tolower(propulsion_code)) %>%
  group_by(generic_make_model) %>%
  mutate(propulsion_code = ifelse(propulsion_code == "undefined", 
                                  first(na.omit(propulsion_code)), 
                                  propulsion_code)) %>%
  ungroup() %>%
  group_by(generic_make_model) %>%
  mutate(engine_capacity_cc = ifelse(is.na(engine_capacity_cc), 
                                     first(na.omit(engine_capacity_cc)), 
                                     engine_capacity_cc)) %>%
  ungroup() %>%
  mutate(propulsion_code = ifelse(propulsion_code == "heavy oil", "diesel", propulsion_code)) %>%
  mutate(propulsion_code = ifelse(propulsion_code == "hybrid electric", "petrol/electric", propulsion_code)) %>%
  filter(generic_make_model != "Data missing or out of range")


#join vehicles_all dan veh_char
vehicles_joined1 <- left_join(vehicles_selection, veh_char,
                            by = c("generic_make_model", "propulsion_code", "engine_capacity_cc"))
vehicles_joined1_succeed <- vehicles_joined1 %>% drop_na(Mass_kg)

#select failed join due to different character but similar potential meaning 
vehicles_failed_rows <- vehicles_joined1 %>%
  filter(is.na(Mass_kg)) %>%
  mutate(model_key = str_extract(generic_make_model, "[A-Z\\s]+\\d*")) %>%
  dplyr::select(-Mass_kg, -SteeringAxle, -Wheelbase, -OtherAxle, -Engine_power)

veh_char_norm <- veh_char %>%
  mutate(model_key = str_extract(generic_make_model, "[A-Z\\s]+\\d*"))

veh_char_norm_unique <- veh_char_norm %>%
  group_by(model_key, propulsion_code, engine_capacity_cc) %>%
  slice(1) %>%
  ungroup()

vehicles_joined2 <- left_join(vehicles_failed_rows, veh_char_norm_unique,
                           by = c("model_key", "propulsion_code", "engine_capacity_cc"),
                           suffix = c("", ".new")) %>%
                           dplyr::select(-model_key, -generic_make_model.new)

vehicles_joined2_succeed <- vehicles_joined2 %>% drop_na(Mass_kg)

vehicles_failed2_rows <- vehicles_joined2 %>%
  filter(is.na(Mass_kg)) %>%
  dplyr::select(-Mass_kg, -SteeringAxle, -Wheelbase, -OtherAxle, -Engine_power)
                           
vehicles_joined3_succeed <- left_join(vehicles_failed2_rows, veh_char1,
                                      by = c("generic_make_model")) %>%
  drop_na(Mass_kg)

vehicles_all_char <- bind_rows(vehicles_joined1_succeed, vehicles_joined2_succeed, vehicles_joined3_succeed)

#joining column vehicle_category on vehicles_all_char to get vehicle class and rating VRU
vehicles_all_char <- left_join(vehicles_all_char, category_vehicles, by = "generic_make_model")

#filter casualties only 0 (pedestrian) and 1 (cyclist)
casualties_filtered <- casualties_selection %>% filter(casualty_type %in% c("Pedestrian", "Cyclist"))

#join data casualty and vehicles based on accident index+vehicle reference
casualties_vehicle <- left_join(casualties_filtered, vehicles_all_char,
                              by = c("collision_index", "vehicle_reference"))

#join data casualty_vehicle and collision to get speed limit as predictor
veh_cas_coll <- left_join(casualties_vehicle, collisions_selection,
                          by = c("collision_index"))
veh_cas_coll <- veh_cas_coll %>% drop_na(generic_make_model)
veh_cas_coll <- veh_cas_coll %>% filter(vehicle_class != "Truck")

#simplified weather variable
veh_cas_coll$weather <- case_when(
  veh_cas_coll$weather_conditions %in% c("Fine no high winds", "Fine + high winds") ~ "Fine",
  veh_cas_coll$weather_conditions %in% c("Raining no high winds", "Raining + high winds") ~ "Raining",
  veh_cas_coll$weather_conditions %in% c("Snowing no high winds", "Snowing + high winds") ~ "Snowing",
  veh_cas_coll$weather_conditions == "Fog or mist" ~ "Fog",
  veh_cas_coll$weather_conditions %in% c("Other", "Unknown") ~ NA
)

#simplified road_surface_conditions variable
veh_cas_coll$road_surface <-veh_cas_coll$road_surface_conditions
veh_cas_coll$road_surface <- case_when(
  veh_cas_coll$road_surface %in% c("Data missing or out of range", "unknown (self reported)") ~ NA,
  TRUE ~ veh_cas_coll$road_surface
)

veh_cas_coll$speed_limit <- case_when(
  veh_cas_coll$speed_limit %in% c("Data missing or out of range") ~ NA,
  TRUE ~ veh_cas_coll$speed_limit
)

#convert to factor in dependent variable (y)
veh_cas_coll$casualty_severity <- factor(veh_cas_coll$casualty_severity, 
                                           levels = c("Slight", "Serious", "Fatal"), 
                                           ordered = TRUE)


#convert to factor in independent variable (x)
veh_cas_coll$speed_limit <- factor(veh_cas_coll$speed_limit,
                                   levels = c("20","30", "40", "50", "60", "70"),
                                   ordered = FALSE)
veh_cas_coll$weather <- factor(veh_cas_coll$weather)
veh_cas_coll$road_surface <- factor(veh_cas_coll$road_surface)
veh_cas_coll$vehicle_class <- factor(veh_cas_coll$vehicle_class)

#scalling data independent variable
veh_cas_coll$Mass_kg10   <- veh_cas_coll$Mass_kg / 10
veh_cas_coll$SteeringAxle10 <- veh_cas_coll$SteeringAxle / 10
veh_cas_coll$Wheelbase10 <- veh_cas_coll$Wheelbase / 10

veh_cas_coll$Mass_kg100   <- veh_cas_coll$Mass_kg / 100
veh_cas_coll$SteeringAxle100 <- veh_cas_coll$SteeringAxle / 100
veh_cas_coll$Wheelbase100 <- veh_cas_coll$Wheelbase / 100


veh_cas_coll$Mass_kg250 <- veh_cas_coll$Mass_kg / 250
veh_cas_coll$engine_capacity100 <- veh_cas_coll$engine_capacity_cc / 100


veh_cas_coll <- veh_cas_coll %>%
  mutate(age_of_vehicle = ifelse(age_of_vehicle == -1, NA, age_of_vehicle))


```

# Initial analysis
Initial analysis was conducted to obtain appropriate statistical testing. Firstly, ordinal logistic regression was considered to use with some reason:
1.Dependent variable that will be examined is casualty severity which has categorical data.
2.The severity consists of slight, serious, and fatal.
3.This category will be changed to factor in the following step to rank based on lower impact to the highest

```{r}
count_casualty <- veh_cas_coll %>%
  count(casualty_severity) %>%
  mutate(prop = round(n / sum(n) * 100, 2))

#proportional odds in model
model_olr1 <- polr(casualty_severity ~ vehicle_class + Rating_VRU + speed_limit + Mass_kg,
                  data = veh_cas_coll,
                  Hess = TRUE)

brant(model_olr1)

```
The amount of data used, particularly casualty_severity in Fatal Condition was very low (1.57%) will may suffer on instability and unreliable data.
Furthermore, result from brant-test (proportional odds test) indicated violations of the proportional odds assumption for several predictors, implying that the effect of these variables is not consistent across the severity levels. which means that increased probability from slight to serious differ with serious to fatal.
Given this limitation, further analysis will merge serious injury and fatal categories into single group labelled as killed/serious injury  (KSI).
then using binary logistic regression in statistical testing.
Consequently, binary logistic regression is applied for subsequent statistical analysis. 
In addition to this consideration, this study also aligns with the Vision Zero strategy adopted in the UK, which aims to eliminate all traffic fatalities and serious injuries. Under this framework, fatal and serious injuries are considered a unified target group, without distinction between them.


```{r}
#adjustment on casualty_severity become binary slight = 0 and serious/fatal= 1
veh_cas_coll$ksi_severity <- ifelse(veh_cas_coll$casualty_severity == "Slight", 0, 1)
```

# Exploratory data analysis
This stage was conducted to explore about type of data and spread of data of independent variable 
```{r}
#check proportion of categorical data

# 1. speed limit category
veh_cas_coll %>%
  count(speed_limit) %>%
  mutate(percent = n/sum(n)*100)

#combine group >=40
veh_cas_coll <- veh_cas_coll %>%
  mutate(speed_limit = as.numeric(as.character(speed_limit)),
         speed_group = case_when(
           speed_limit == 20 ~ "Low",
           speed_limit == 30 ~ "Medium",
           speed_limit >= 40 ~ "High",
           TRUE ~ NA_character_
         )) %>%
  mutate(speed_group = factor(speed_group, levels = c("Low", "Medium", "High")))

#check distribution based on KSI severity (dependent category)
veh_cas_coll %>%
  group_by(speed_group,ksi_severity) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

# 2. road surface
veh_cas_coll %>%
  count(road_surface) %>%
  mutate(percent = n/sum(n)*100)

#combine group, become binary dry vs non dry
veh_cas_coll <- veh_cas_coll %>%
  mutate(road_surface_group = case_when(
    road_surface == "Dry" ~ "Dry",
    road_surface %in% c("Wet or damp", "Frost or ice", "Snow", "Flood over 3cm. deep") ~ "Non-dry",
    TRUE ~ NA_character_
  )) %>%
  mutate(road_surface_group = factor(road_surface_group, levels = c("Dry", "Non-dry")))

#check distribution based on KSI severity (dependent category)
veh_cas_coll %>%
  group_by(road_surface_group,ksi_severity) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

# 3. weather
veh_cas_coll %>%
  count(weather) %>%
  mutate(percent = n/sum(n)*100)

#combine group, become binary Fine vs not fine
veh_cas_coll <- veh_cas_coll %>%
  mutate(weather_group = case_when(
    weather == "Fine" ~ "Fine",
    weather %in% c("Raining", "Fog", "Snowing") ~ "Not fine",
    TRUE ~ NA_character_
  )) %>%
  mutate(weather_group = factor(weather_group, levels = c("Fine", "Not fine")))

#check distribution based on KSI severity (dependent category)
veh_cas_coll %>%
  group_by(weather_group,ksi_severity) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

# 4. light conditions
veh_cas_coll %>%
  count(light_conditions) %>%
  mutate(percent = n/sum(n)*100)

#combine category, become daylight vs darkness,
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    light_group = case_when(
      light_conditions == "Daylight" ~ "Daylight",
      light_conditions %in% c("Darkness - lights lit","Darkness - no lighting", "Darkness - lights unlit") ~ "Darkness",
      light_conditions %in% c(
        "Data missing or out of range", 
        "Darkness - lighting unknown"
      ) ~ NA_character_,
      TRUE ~ NA_character_
    )
  )

#check distribution based on KSI severity (dependent category)
veh_cas_coll %>%
  group_by(light_group,ksi_severity) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

# 5. first point of impact
veh_cas_coll %>%
  count(first_point_of_impact) %>%
  mutate(percent = n/sum(n)*100)

#delete other useless category,
veh_cas_coll <- veh_cas_coll %>%
  mutate(impact_group = case_when(
    first_point_of_impact == "Front" ~ "Front",
    first_point_of_impact == "Nearside" ~ "Nearside",
    first_point_of_impact == "Offside" ~ "Offside",
    first_point_of_impact == "Back" ~ "Back",
    TRUE ~ NA_character_
  )) %>%
  mutate(impact_group = factor(impact_group, levels = c("Front", "Nearside", "Offside", "Back")))

#check distribution based on KSI severity (dependent category)
veh_cas_coll %>%
  group_by(impact_group,ksi_severity) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

# 6. urban / rural area
veh_cas_coll %>%
  count(urban_or_rural_area) %>%
  mutate(percent = n/sum(n)*100)

#define category into urban or rural only,
veh_cas_coll <- veh_cas_coll %>%
  mutate(urban_rural = ifelse(
    urban_or_rural_area %in% c("Unallocated", "Data missing or out of range"),
    NA,
    urban_or_rural_area
  ))

# check outlier of independent variable with continuous data
# 7. Mass vehicle
ggplot(veh_cas_coll, aes(y = Mass_kg250)) +
  geom_boxplot() +
  theme_minimal()

#replace outlier of Mass_kg with NA
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    Mass_kg_raw = Mass_kg250,
    Mass_kg250 = ifelse(
      Mass_kg250 < quantile(Mass_kg_raw, 0.25, na.rm = TRUE) - 1.5 * IQR(Mass_kg_raw, na.rm = TRUE) |
        Mass_kg250 > quantile(Mass_kg_raw, 0.75, na.rm = TRUE) + 1.5 * IQR(Mass_kg_raw, na.rm = TRUE),
      NA,
      Mass_kg_raw
    )
  )

# 8. Rating VRU
ggplot(veh_cas_coll, aes(y = Rating_VRU)) +
  geom_boxplot() +
  theme_minimal()

#replace outlier of Rating VRU with NA
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    Rating_VRU_raw = Rating_VRU,
    Rating_VRU = ifelse(
      Rating_VRU < quantile(Rating_VRU_raw, 0.25, na.rm = TRUE) - 1.5 * IQR(Rating_VRU_raw, na.rm = TRUE) |
        Rating_VRU > quantile(Rating_VRU_raw, 0.75, na.rm = TRUE) + 1.5 * IQR(Rating_VRU_raw, na.rm = TRUE),
      NA,
      Rating_VRU_raw
    )
  )

# 9. Engine_power
ggplot(veh_cas_coll, aes(y = Engine_power)) +
  geom_boxplot() +
  theme_minimal()

#replace outlier of Engine_power with NA
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    Engine_power_raw = Engine_power,
    Engine_power = ifelse(
      Engine_power < quantile(Engine_power_raw, 0.25, na.rm = TRUE) - 1.5 * IQR(Engine_power_raw, na.rm = TRUE) |
        Engine_power > quantile(Engine_power_raw, 0.75, na.rm = TRUE) + 1.5 * IQR(Engine_power_raw, na.rm = TRUE),
      NA,
      Engine_power_raw
    )
  )

# 10. engine_capacity_cc
ggplot(veh_cas_coll, aes(y = engine_capacity100)) +
  geom_boxplot() +
  theme_minimal()

#replace outlier of engine_capacity_cc with NA
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    engine_capacity_cc_raw = engine_capacity100,
    engine_capacity100 = ifelse(
      engine_capacity100 < quantile(engine_capacity_cc_raw, 0.25, na.rm = TRUE) - 1.5 * IQR(engine_capacity_cc_raw, na.rm = TRUE) |
        engine_capacity100 > quantile(engine_capacity_cc_raw, 0.75, na.rm = TRUE) + 1.5 * IQR(engine_capacity_cc_raw, na.rm = TRUE),
      NA,
      engine_capacity_cc_raw
    )
  )

# 11. age of vehicle
ggplot(veh_cas_coll, aes(y = age_of_vehicle)) +
  geom_boxplot() +
  theme_minimal()

#replace outlier of age of vehicle with NA
veh_cas_coll <- veh_cas_coll %>%
  mutate(
    age_of_vehicle_raw = age_of_vehicle,
    age_of_vehicle = ifelse(
      age_of_vehicle < quantile(age_of_vehicle_raw, 0.25, na.rm = TRUE) - 1.5 * IQR(age_of_vehicle_raw, na.rm = TRUE) |
        age_of_vehicle > quantile(age_of_vehicle_raw, 0.75, na.rm = TRUE) + 1.5 * IQR(age_of_vehicle_raw, na.rm = TRUE),
      NA,
      age_of_vehicle_raw
    )
  )

# 12. age_of_casualty
ggplot(veh_cas_coll, aes(y = age_of_casualty)) +
  geom_boxplot() +
  theme_minimal()

# Checking NA values in each variable
na_summary <- veh_cas_coll %>%
  summarise(
    speed_group         = sum(is.na(speed_group)),
    road_surface_group  = sum(is.na(road_surface_group)),
    weather_group       = sum(is.na(weather_group)),
    light_group         = sum(is.na(light_group)),
    impact_group        = sum(is.na(impact_group)),
    urban_rural         = sum(is.na(urban_rural)),
    Mass_kg             = sum(is.na(Mass_kg)),
    Rating_VRU          = sum(is.na(Rating_VRU)),
    Engine_power        = sum(is.na(Engine_power)),
    engine_capacity_cc  = sum(is.na(engine_capacity_cc)),
    age_of_vehicle      = sum(is.na(age_of_vehicle)),
    age_of_casualty     = sum(is.na(age_of_casualty))
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "na_count") %>%
  mutate(
    total_rows = nrow(veh_cas_coll),
    percent_na = round(100 * na_count / total_rows, 2)
  ) %>%
  dplyr::select(variable, na_count, percent_na)


```

```{r}
# decision independent variable : not using age_of_vehicle and engine_power due to higher proportion of NA (> 20%)
veh_cas_clean <- veh_cas_coll %>% drop_na(
  Rating_VRU, Mass_kg250, engine_capacity100,
  age_of_casualty,
  speed_group, light_group, weather_group, road_surface_group,impact_group,urban_rural
)
```


# Determine Model Variable
Assessment is conducted to understand the influence of vehicle and road characteristics.
as consequently, model is built based on theory some of factors influencing casualty severity among pedestrian and cyclists as follows:
Model 2 - Vehicle + Road feature + Confounding Factors
Model was built as influence of that features
1. Model 0 - confounding factors
2. Model 1 - Model 0 + Vehicle -> Confounding + vehicle
3. Model 2 - Model 1 + Road -> Confounding + vehicle + road


```{r}
#Model 0 - multivariate binary logistic variable (confounding factor)
model_multi_con <- glm(
  ksi_severity ~  age_of_casualty +
                  light_group+weather_group+road_surface_group+impact_group+urban_rural,
  data = veh_cas_clean,
  family = "binomial"
)

exp(coef(model_multi_con)) 

summary(model_multi_con)

#evaluate model using pseudo R-squared
pR2(model_multi_con)

#evaluation of model 0
AIC(model_multi_con)

#evaluation on AUC indicator
prob1 <- predict(model_multi_con, newdata = veh_cas_clean, type = "response")
valid_idx <- !is.na(prob1)
pred <- prediction(prob1[valid_idx], veh_cas_clean$ksi_severity[valid_idx])
perf <- performance(pred, "auc")
auc_value1 <- as.numeric(perf@y.values)
print(auc_value1)

#plot odds ratio
model_summary0 <- summary(model_multi_con)
coef_table0 <- model_summary0$coefficients

num_predictor0 <- length(coef(model_multi_con))  
coef_table0 <- coef_table0[1:num_predictor0, , drop = FALSE]

# calculation of Odds Ratio
estimates_OR0   <- coef_table0[,"Estimate"]
std_error0      <- coef_table0[,"Std. Error"]
Odds_Ratio0     <- exp(estimates_OR0)
lower_CI0  <- exp(estimates_OR0 - 1.96 * std_error0)
upper_CI0  <- exp(estimates_OR0 + 1.96 * std_error0)
  
# create dataframe for ploting to ggplot
plot_df0 <- data.frame(
  variable    = rownames(coef_table0),
  Odds_Ratio  = Odds_Ratio0,
  lower       = lower_CI0,
  upper       = upper_CI0
)

plot_df0$variable <- factor(plot_df0$variable, 
                             levels = rownames(coef_table0))

var_labels0 <- c(
  "(Intercept)" = "Intercept",
  "age_of_casualty" = "Age of Casualty",
  "light_groupDaylight" = "Daylight",
  "weather_groupNot fine" = "Weather: Not Fine",
  "road_surface_groupNon-dry" = "Road Surface: Non-dry",
  "impact_groupNearside" = "Impact: Nearside",
  "impact_groupOffside" = "Impact: Offside",
  "impact_groupBack" = "Impact: Back",
  "urban_ruralUrban" = "Urban Area"
)

# Ganti kolom variable dengan label
plot_df0$variable <- recode(plot_df0$variable, !!!var_labels)

# Plot ulang
ggplot(plot_df0, aes(x = Odds_Ratio, y = variable)) +
  geom_point(size = 2) +  # titik lebih besar
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25, size = 1) + # error bar lebih tebal
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) + # garis merah lebih tebal
  scale_x_log10() +
  labs(
    title = "Odds Ratio Plot (Multivariate Binary Logistic Regression)",
    x = "Odds Ratio",
    y = "Predictors"
  ) +
  theme_minimal(base_size = 14) + # ukuran font dasar lebih besar
  theme(
    axis.text = element_text(color = "black", face = "bold"),   # teks sumbu lebih tebal
    axis.title = element_text(color = "black", face = "bold"),  # judul sumbu lebih tebal
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16) # judul ke tengah, tebal
  )

```

```{r}
#Model 1 - multivariate binary logistic variable (confounding factor + vehicle)
model_multi_veh <- glm(
  ksi_severity ~  Rating_VRU+ Mass_kg250+engine_capacity100+
                  age_of_casualty +
                  light_group+weather_group+road_surface_group+impact_group+urban_rural,
  data = veh_cas_clean,
  family = "binomial"
)

exp(coef(model_multi_veh)) 

summary(model_multi_veh)

#evaluate model using pseudo R-squared
pR2(model_multi_veh)

#evaluation of model 2 to compare with another model
AIC(model_multi_veh)

#evaluation on AUC indicator
prob2 <- predict(model_multi_veh, newdata = veh_cas_clean, type = "response")
valid_idx <- !is.na(prob2)
pred <- prediction(prob2[valid_idx], veh_cas_clean$ksi_severity[valid_idx])
perf <- performance(pred, "auc")
auc_value2 <- as.numeric(perf@y.values)
print(auc_value2)


#plot odds ratio
model_summary1 <- summary(model_multi_veh)
coef_table1 <- model_summary1$coefficients

num_predictor1 <- length(coef(model_multi_veh))  
coef_table1 <- coef_table1[1:num_predictor1, , drop = FALSE]

# calculation of Odds Ratio
estimates_OR1   <- coef_table1[,"Estimate"]
std_error1      <- coef_table1[,"Std. Error"]
Odds_Ratio1     <- exp(estimates_OR1)
lower_CI1  <- exp(estimates_OR1 - 1.96 * std_error1)
upper_CI1  <- exp(estimates_OR1 + 1.96 * std_error1)
  
# create dataframe for ploting to ggplot
plot_df1 <- data.frame(
  variable    = rownames(coef_table1),
  Odds_Ratio  = Odds_Ratio1,
  lower       = lower_CI1,
  upper       = upper_CI1
)

plot_df1$variable <- factor(plot_df1$variable, 
                             levels = rownames(coef_table1))

var_labels <- c(
  "(Intercept)" = "Intercept",
  "Rating_VRU" = "VRU Safety Rating",
  "Mass_kg250" = "Vehicle Weight (per 250 kg)",
  "engine_capacity100" = "Engine Capacity (per 100 cc)",
  "age_of_casualty" = "Age of Casualty",
  "light_groupDaylight" = "Daylight",
  "weather_groupNot fine" = "Weather: Not Fine",
  "road_surface_groupNon-dry" = "Road Surface: Non-dry",
  "impact_groupNearside" = "Impact: Nearside",
  "impact_groupOffside" = "Impact: Offside",
  "impact_groupBack" = "Impact: Back",
  "urban_ruralUrban" = "Urban Area"
)

# Ganti kolom variable dengan label
plot_df1$variable <- recode(plot_df1$variable, !!!var_labels)

# Plot ulang
ggplot(plot_df1, aes(x = Odds_Ratio, y = variable)) +
  geom_point(size = 2) +  # titik lebih besar
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25, size = 1) + # error bar lebih tebal
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) + # garis merah lebih tebal
  scale_x_log10() +
  labs(
    title = "Odds Ratio Plot (Multivariate Binary Logistic Regression)",
    x = "Odds Ratio",
    y = "Predictors"
  ) +
  theme_minimal(base_size = 14) + # ukuran font dasar lebih besar
  theme(
    axis.text = element_text(color = "black", face = "bold"),   # teks sumbu lebih tebal
    axis.title = element_text(color = "black", face = "bold"),  # judul sumbu lebih tebal
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16) # judul ke tengah, tebal
  )
```

```{r}
#Model 2 - multivariate binary logistic variable (confounding factor + vehicle + environment)
model_multi_veh_road <- glm(
  ksi_severity ~  Rating_VRU+ Mass_kg250+engine_capacity100+
                  speed_group+
                  age_of_casualty +
                  light_group+weather_group+road_surface_group+impact_group+urban_rural,,
  data = veh_cas_clean,
  family = "binomial"
)

exp(coef(model_multi_veh_road)) 

summary(model_multi_veh_road)

#evaluate model using pseudo R-squared
pR2(model_multi_veh_road)

#evaluation of model 3 to compare with another model
AIC(model_multi_veh_road)

#evaluation on AUC indicator
prob3 <- predict(model_multi_veh_road, newdata = veh_cas_clean, type = "response")
valid_idx <- !is.na(prob3)
pred <- prediction(prob3[valid_idx], veh_cas_clean$ksi_severity[valid_idx])
perf <- performance(pred, "auc")
auc_value3 <- as.numeric(perf@y.values)
print(auc_value3)
```

#Model Performance
contribution of adding independent variable
```{r}
pR2_values <- c(
  model0 = pR2(model_multi_con)["McFadden"],
  model1 = pR2(model_multi_veh)["McFadden"],
  model2 = pR2(model_multi_veh_road)["McFadden"]
)

AIC_values <- c(
  model0 = AIC(model_multi_con),
  model1 = AIC(model_multi_veh),
  model2 = AIC(model_multi_veh_road)
)

AUC_value <- c(
  model0 = auc_value1,
  model1 = auc_value2,
  model2 = auc_value3
)

comparison_table <- data.frame(
  Model_0 = c(pR2(model_multi_con)["McFadden"], AIC(model_multi_con), auc_value1),
  Model_1 = c(pR2(model_multi_veh)["McFadden"], AIC(model_multi_veh), auc_value2),
  Model_2 = c(pR2(model_multi_veh_road)["McFadden"], AIC(model_multi_veh_road), auc_value3)
)

rownames(comparison_table) <- c("McFadden R²", "AIC", "AUC")

print(comparison_table)

```

Model Diagnostics
```{r}
# Convergence
model_multi_veh_road$converged

# multi collinearity
vif_values <- vif(model_multi_veh_road)

# Ubah menjadi dataframe
vif_df <- data.frame(
  Variable = names(vif_values),
  VIF = as.numeric(vif_values),
  row.names = NULL
)
vif_df
```

#Model Interpretation
visualization:
```{r}
#plot odds ratio
model_summary <- summary(model_multi_veh_road)
coef_table <- model_summary$coefficients

num_predictor <- length(coef(model_multi_veh_road))  
coef_table <- coef_table[1:num_predictor, , drop = FALSE]

# calculation of Odds Ratio
estimates_OR   <- coef_table[,"Estimate"]
std_error      <- coef_table[,"Std. Error"]
Odds_Ratio     <- exp(estimates_OR)
lower_CI  <- exp(estimates_OR - 1.96 * std_error)
upper_CI  <- exp(estimates_OR + 1.96 * std_error)
  
# create dataframe for ploting to ggplot
plot_df <- data.frame(
  variable    = rownames(coef_table),
  Odds_Ratio  = Odds_Ratio,
  lower       = lower_CI,
  upper       = upper_CI
)

plot_df$variable <- factor(plot_df$variable, 
                             levels = rownames(coef_table))

var_labels <- c(
  "(Intercept)" = "Intercept",
  "Rating_VRU" = "VRU Safety Rating",
  "Mass_kg250" = "Vehicle Weight (per 250 kg)",
  "engine_capacity100" = "Engine Capacity (per 100 cc)",
  "speed_groupMedium" = "Speed Limit: Medium",
  "speed_groupHigh" = "Speed Limit: High",
  "age_of_casualty" = "Age of Casualty",
  "light_groupDaylight" = "Daylight",
  "weather_groupNot fine" = "Weather: Not Fine",
  "road_surface_groupNon-dry" = "Road Surface: Non-dry",
  "impact_groupNearside" = "Impact: Nearside",
  "impact_groupOffside" = "Impact: Offside",
  "impact_groupBack" = "Impact: Back",
  "urban_ruralUrban" = "Urban Area"
)

# Ganti kolom variable dengan label
plot_df$variable <- recode(plot_df$variable, !!!var_labels)

# Plot ulang
ggplot(plot_df, aes(x = Odds_Ratio, y = variable)) +
  geom_point(size = 2) +  # titik lebih besar
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25, size = 1) + # error bar lebih tebal
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 1) + # garis merah lebih tebal
  scale_x_log10() +
  labs(
    title = "Odds Ratio Plot (Multivariate Binary Logistic Regression)",
    x = "Odds Ratio",
    y = "Predictors"
  ) +
  theme_minimal(base_size = 14) + # ukuran font dasar lebih besar
  theme(
    axis.text = element_text(color = "black", face = "bold"),   # teks sumbu lebih tebal
    axis.title = element_text(color = "black", face = "bold"),  # judul sumbu lebih tebal
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16) # judul ke tengah, tebal
  )
```

convert to probability of each variable
```{r}
#Speed limit
# AME for categorical variable
ame_speed <- avg_comparisons(
  model_multi_veh_road,
  variables = "speed_group",
  vcov = "HC3"
)
summary(ame_speed)

ame_med_high <- avg_comparisons(
  model_multi_veh_road,
  variables = list(speed_group = c("Medium", "High")),
  vcov = "HC3"
)
summary(ame_med_high)

eff_speed <- ggpredict(model_multi_veh_road, terms = "speed_group")
pred_df <- data.frame(
  speed_group = eff_speed$x,
  prob = eff_speed$predicted,
  conf.low = eff_speed$conf.low,
  conf.high = eff_speed$conf.high
  )

ame_df <- data.frame(
  from = c("Low", "Low", "Medium"),
  to   = c("Medium", "High", "High"),
  ame  = c(
    ame_speed$estimate[ame_speed$contrast == "Medium - Low"],
    ame_speed$estimate[ame_speed$contrast == "High - Low"],
    ame_med_high$estimate[ame_med_high$contrast == "High - Medium"]
  )
)
pred_df$speed_group <- factor(pred_df$speed_group,
                              levels = c("High", "Low", "Medium"))

ame_plot <- ame_df %>%
  left_join(pred_df, by = c("from" = "speed_group")) %>%
  rename(prob_from = prob,
         conf.low_from = conf.low,
         conf.high_from = conf.high) %>%
  left_join(pred_df, by = c("to" = "speed_group")) %>%
  rename(prob_to = prob,
         conf.low_to = conf.low,
         conf.high_to = conf.high)

ame_plot <- ame_plot %>%
  mutate(
    xmid = (as.numeric(from) + as.numeric(to)) / 2,
    ymid = (prob_from + (prob_from + ame)) / 2
  )

ggplot(pred_df, aes(x = speed_group, y = prob)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, size=1) +
  geom_point(size = 4, color = "steelblue") +
  # Panah AME
  geom_curve(data = ame_plot,
             aes(x = from, xend = to,
                 y = prob_from,
                 yend = prob_from + ame),
             curvature = 0.2, arrow = arrow(length = unit(0.02, "npc")),
             color = "darkred", inherit.aes = FALSE, size=1) +
  # Label AME
  geom_text(data = ame_plot,
          aes(x = xmid,
              y = ymid,
              label = paste0("+", round(ame*100, 1), " pp")),
          vjust = -0.5, color = "darkred", inherit.aes = FALSE)+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Speed Limit Category",
    y = "Predicted Probability of KSI",
    title = "Predicted Probability & Marginal Effects of Speed Limit",
  ) +
  theme_minimal(base_size = 14)+
  theme(
    plot.title    = element_text(hjust = 0.5, face = "bold", size = 16, color = "black"),
    axis.title    = element_text(face = "bold", size = 14, color = "black"),
    axis.text     = element_text(size = 12, color = "black"),
    panel.grid    = element_line(color = "grey80")  # grid tetap abu2 biar lembut
  )



```

```{r}
# average marginal effects for continous variables
ame_rating <- avg_slopes(
  model_multi_veh_road,
  variables = c("Rating_VRU"),
  vcov = "HC3"  # robust SE
)
summary(ame_rating)
# VRU Rating
eff_rating <- ggpredict(model_multi_veh_road, terms = "Rating_VRU")
plot(eff_rating) +
  labs(
    x = "VRU Safety Rating",
    y = "Predicted Probability of KSI",
    title = "Predicted Probability of KSI on Vehicle Rating",
  )+
  theme(
    plot.title    = element_text(hjust = 0.5, face = "bold", size = 16, color = "black"),
    axis.title    = element_text(face = "bold", size = 14, color = "black"),
    axis.text     = element_text(size = 12, color = "black"),
    panel.grid    = element_line(color = "grey80")  # grid tetap abu2 biar lembut
  )
  
```

#weighted model based for RQ3
```{r}
p_speed = max(ame_df$ame)
p_vru = abs(ame_rating$estimate)

#weighted model based each variable
w_speed = p_speed/(p_speed+p_vru)
w_vru = p_vru/(p_speed+p_vru)

```